---
title: "May Institute 2017 - Program 5 : Multivariate analysis and classification"
author: "Meena Choi"
date: "5/10/2017"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
---
```{r, echo=F, warning=F}
## we're setting up the directory for figures
library(knitr)
opts_chunk$set(fig.path = "./")
opts_chunk$set(fig.align = "center")
```

***

# Data
- the quantified peak intensities data from ABRF 2015, processed by Skyline.
- The processed data, `quant.skyline.rda`, which is output of `dataProcess` from [section 2](https://github.com/MayInstitute/MayInstitute2017/blob/master/Program5_Stat%20for%20MS/MayInst_P5_sec2.html) 

***

# 1. Protein quantification from the output of `dataProcess` in MSstats

## 1.1 Read the output of `dataProcess` 
```{r}
load('./data_day3/quant.skyline.rda')
```

It should use the output of `quantification` function in MSstats in order to get subject quantification for each protein. In this ABRF study, there is no biological replicate. Let's use run-level quantification for protein instead as an example.

```{r}
head(quant.skyline$RunlevelData)
runquant <- quant.skyline$RunlevelData
head(runquant)
```

## 1.2 Reformat to wide format
Then let's make wide format of data from long format.
```{r}
library(reshape2)

input.wide <- dcast(Protein ~ originalRUN, data=runquant, value.var = 'LogIntensities')
head(input.wide)
rownames(input.wide) <- input.wide$Protein
input.wide <- input.wide[, -1]
head(input.wide)
dim(input.wide) # there are 3027 rows (one row per protein) and 12 columns (each colomn for run)
```

Please check whether there are missing values (NAs) or not. If there is no intensity at all in certain run for certain protein, we can't get run-summarized for that run.
```{r}
sum(is.na(input.wide))
```

There are three NAs. Then, we need to decide how to deal with NAs.
* First option: remove the samples with missing values
* Second option: impute the missing values. Then need to how to impute. 1) impute with zero, 2) with minimum value per protein or among all proteins, 3) with median or mean, 4) random sampling among other run-summarized value per protein.

In this case, let's remove proteins with any NAs.
```{r}
samplemissing <- apply(input.wide, 1, function(x) sum(is.na(x)))
unique(samplemissing)
selectedSamples <- which(samplemissing == 0) 
# Only keep the samples with no missing values
input.wide <- input.wide[selectedSamples, ]
dim(input.wide)
```
Two samples, which have any NAs, are removed.

Let's make another data with only spike-in proteins.
```{r}
input.wide.spike <- input.wide[rownames(input.wide) %in% c("sp|P44015|VAC2_YEAST",
                                            "sp|P55752|ISCB_YEAST",
                                            "sp|P44374|SFG2_YEAST",
                                            "sp|P44983|UTR6_YEAST",
                                            "sp|P44683|PGA4_YEAST",
                                            "sp|P55249|ZRT4_YEAST"),]

input.wide.spike
```

I will save two wide format data.
```{r}
save(input.wide.spike, file = 'input.wide.spike.rda')
save(input.wide.spike, file = 'input.wide.spike.rda')
```


# 2. Perform principal components analysis (PCA) on the given data matrix

## 2.1 Only 6 spike-in proteins

### 2.1.1 PCA with `prcomp` function

First, let's try PCA with 6 spike-in proteins only. Input has the row for proteins and the column for run (subject).

```{r}
head(input.wide.spike)
pc <- prcomp(t(input.wide.spike))

# Inspect PCA object
summary(pc)
names(pc)
```

### 2.1.2 Check the proportion of explained variance

Let's check the proportion of explained variance. The first component has the largest variance. In this case, we need 3 components to capture most of the variation.
```{r}
percent_var <- pc$sdev^2/sum(pc$sdev^2)
barplot(percent_var, xlab="Principle component", ylab="% of variance")

cum_var <- cumsum(pc$sdev^2/sum(pc$sdev^2))
barplot(cum_var, xlab="Principle component", ylab="Cumulative % of variance" )
```

### 2.1.3 Visualization for PC1 vs PC2

Let's visualize PC1 vs PC2 in scatterplot. 'x' include PC components for each subject.

```{r}
head(pc$x)

library(ggplot2)

ggplot(aes(x=PC1, y=PC2), data=data.frame(pc$x))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()
```

In order to distinguish group by colors or shape, create vector of condition labels variable called Condition. The order should be the same as column of input.

```{r}
colnames(input.wide.spike)
Condition <- c(rep("Condition1", 3), rep("Condition2", 3),
               rep("Condition3", 3), rep("Condition4", 3))
Condition

# Create PC1 vs PC2 scatterplot with Condition colors
ggplot(aes(x=PC1, y=PC2, color=Group), data=data.frame(pc$x, Group=Condition))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()
```

We can create PCA scatterplots with custom Condition labels and colors.

```{r}
ggplot(aes(x=PC1, y=PC2, color=Group, shape=Group), data=data.frame(pc$x, Group=Condition))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()

ggplot(aes(x=PC1, y=PC3, color=Group, shape=Group), data=data.frame(pc$x, Group=Condition))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()

ggplot(aes(x=PC2, y=PC3, color=Group, shape=Group), data=data.frame(pc$x, Group=Condition))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()
```

### 2.1.4 Check the PC1 loadings

Let's create plot of PC1 loadings. 'rotation' show the Protein loading for each component.
```{r}
head(pc$rotation)

ggplot(aes(x=Protein, xend=Protein, y=0, yend=PC1), 
       data=data.frame(pc$rotation, Protein=rownames(pc$rotation)))+
  geom_segment()+
  labs(y="PC1 loading")+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))
```

### 2.1.5 loadings vs p-values

Let's compare to loadings to p-values. `loadings` close to zero mean less contribution. Many proteins with small loadings have significant p-values.

```{r}

load('./data_day3/Skyline.result.rda')
head(Skyline.result)

Skyline.result.spikein <- Skyline.result[Skyline.result$Protein %in% c("sp|P44015|VAC2_YEAST",
                                            "sp|P55752|ISCB_YEAST",
                                            "sp|P44374|SFG2_YEAST",
                                            "sp|P44983|UTR6_YEAST",
                                            "sp|P44683|PGA4_YEAST",
                                            "sp|P55249|ZRT4_YEAST"), ]
Skyline.result.spikein

smoothScatter(pc$rotation[,1], Skyline.result.spikein[Skyline.result.spikein$Label == 'C2-C1', 'adj.pvalue'])
smoothScatter(pc$rotation[,1], Skyline.result.spikein[Skyline.result.spikein$Label == 'C3-C1', 'adj.pvalue'])
```


## 2.2 All 3025 proteins

### 2.2.1 PCA with `prcomp` function
Let's repeat with all proteins (3025 proteins).
```{r}
head(input.wide)
dim(input.wide)

pc <- prcomp(t(input.wide))

# Inspect PCA object
summary(pc)
names(pc)
```

### 2.2.2 Check the proportion of explained variance

Let's check the proportion of explained variance. The first component has the largest variance. In this case, we need 10-11 components to capture most of the variation.
```{r}
percent_var <- pc$sdev^2/sum(pc$sdev^2)
barplot(percent_var, xlab="Principle component", ylab="% of variance")

cum_var <- cumsum(pc$sdev^2/sum(pc$sdev^2))
barplot(cum_var, xlab="Principle component", ylab="Cumulative % of variance" )
```

### 2.1.3 Visualization for PC1 vs PC2

Let's visualize PC1 vs PC2 in scatterplot. 'x' include PC components for each subject.

```{r}
# 'x' include PC components for each subject.
head(pc$x)

ggplot(aes(x=PC1, y=PC2), data=data.frame(pc$x))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()
```

In order to make different colors or shapes by groups, create Create vector of disease labels variable called Condition.
```{r}
colnames(input.wide)
Condition <- c(rep("Condition1", 3), rep("Condition2", 3),
               rep("Condition3", 3), rep("Condition4", 3))
Condition

# Create PC1 vs PC2 scatterplot with Condition colors
ggplot(aes(x=PC1, y=PC2, color=Group), data=data.frame(pc$x, Group=Condition))+
  geom_point(size=4, alpha=0.5)+
  theme_bw()
```

With more number proteins, we need more PC components to reach most of variation.

***

# 3. Heatmap

## 3.1 Only 6 spike-in proteins

### 3.1.1 matrix format
```{r}
# check the class
class(input.wide.spike)

# It is data.frame. Convert to numeric matrix
input.wide.spike <- as.matrix(input.wide.spike)
class(input.wide.spike)

# Visually no difference
head(input.wide.spike)
```

### 3.1.2 `heatmap` function in base `stats` package

First, let's try to draw heatmap with base function.
```{r}
heatmap(input.wide.spike, scale='none')
heatmap(input.wide.spike, scale='row')
heatmap(input.wide.spike, scale='column')

# Change the font of row and column label
heatmap(input.wide.spike, cexRow = 0.6, cexCol = 0.6, margins = c(2, 2))
# Don't do cluster on rows
heatmap(input.wide.spike, cexRow = 0.6, cexCol = 0.6, margins = c(2, 2), Rowv = NA)
# Don't do cluster on columns
heatmap(input.wide.spike, cexRow = 0.6, cexCol = 0.6, margins = c(2, 2), Colv = NA)
```

### 3.1.3 Color bar for group information
Add color side bar at th top of columns to distinguish group information by run.

```{r}
group.color <- c(rep("red", 3), rep("orange", 3),
               rep("yellow", 3), rep("blue", 3))
heatmap(input.wide.spike, ColSideColors=group.color)
```

### 3.1.4 Color scale
If color scale changed, it is easy to see the difference.

```{r}
library(marray)
my.colors <- c(maPalette(low = "darkblue", high = "white", k = 7)[-7],
               "white",maPalette(low = "white", high = "darkred", k = 7)[-1])
heatmap(input.wide.spike, 
        col=my.colors, 
        ColSideColors=group.color,
        scale='none')
```

### 3.1.5 Different distance and clustering

Try different distances calculation and clustering methods. Choice of distance metric or clustering matters!

* Distance options: euclidean (default), maximum, canberra, binary, minkowski, manhattan

* Cluster options: complete (default), single, average, mcquitty, median, centroid, ward

```{r}
# can change method for distance calculation
col_distance <- dist(t(input.wide.spike), method = "euclidean")
# can change clustering method
col_cluster <- hclust(col_distance, method = "ward.D")

heatmap(input.wide.spike, 
        col=my.colors, 
        ColSideColors=group.color,
        Colv = as.dendrogram(col_cluster),
        scale='none')
```



## 3.2 All proteins (3025 proteins)

### 3.2.1 matrix format
```{r}
# check the class
class(input.wide)

# It is data.frame. Convert to numeric matrix
input.wide <- as.matrix(input.wide)
class(input.wide)

# Visually no difference
head(input.wide)
```

### 3.2.3 Color bar and non-scaled data

Without scale in rows, it is hard to see the difference between groups per protein.
```{r}
group.color <- c(rep("red", 3), rep("orange", 3),
               rep("yellow", 3), rep("blue", 3))

# remove row information (Protein IDs)
heatmap(input.wide, 
        ColSideColors=group.color,
        col=my.colors,
        labRow = FALSE,
        scale = 'none')
```

Protein quantification will be centered and scaled in the row direction.

```{r}
# no Protein IDs
heatmap(input.wide, 
        ColSideColors=group.color,
        col=my.colors,
        labRow = FALSE,
        scale = 'row')
```

If we change distance calculation and/or clustering method, dendrogram is different. In this case, clustering between methods are not much different.

```{r}
# can change method for distance calculation
col_distance <- dist(t(input.wide), method = "euclidean")
# can change clustering method
col_cluster <- hclust(col_distance, method = "ward.D")

heatmap(input.wide, 
        col=my.colors, 
        ColSideColors=group.color,
        Colv = as.dendrogram(col_cluster),
        labRow = FALSE,
        scale = 'row')
```

***

# 4. Biological data 

## 4.1 Read subject quantification per protein and handle missing values.
In this section we’ll continue using mpg and CRC dataset.

```{r}
crc <- read.csv("./data_day3/CRC_train.csv")
head(crc)
colnames(crc)
dim(crc)

crc.prot <- crc[,1:72]
crc.anno <- crc[,73:79]

## check missing value
sum(is.na(crc.prot))

# Deal with missing value
# First option: remove the samples with missing values
dim(na.omit(crc.prot))

# Second option: impute the missing values
random.imp <- function (a){
  missing <- is.na(a)
  n.missing <- sum(missing)
  a.obs <- a[!missing]
  imputed <- a
  # imputed[missing] <- 0
  # imputed[missing] <- sample(a.obs, n.missing, replace=TRUE)
  # imputed[missing] <- median(a.obs)
  imputed[missing] <- min(a.obs)
  return (imputed)
}

pMiss <- function(x){
    sum(is.na(x))/length(x)*100
}

# 1. Only keep the proteins with less than 5% missing values
proteinmissing <- apply(crc.prot, 2, pMiss)
selectedprotein <- which(proteinmissing <= 5) 
crc.prot <- crc.prot[, selectedprotein]

# 2. Only keep the samples with less than 5% missing values
samplemissing <- apply(crc.prot, 1, pMiss)
selectedSamples <- which(samplemissing <= 5) 
# Then, impute the missing value
imputed.crc.prot <- t(apply(crc.prot[selectedSamples,], 1, function(x) random.imp(x)))
imputed.crc.anno <- crc.anno[selectedSamples, ]
dim(imputed.crc.anno)
```

## 4.2 Heatmap of individual values of proteins

### 4.2.1 Default heatmap
```{r}
library(genefilter)
## heatmap function in base stats package
heatmap(t(imputed.crc.prot))
```

### 4.2.2 Change color

```{r}
# Non-scaled 
heatmap(t(imputed.crc.prot), cexRow = 0.3, cexCol = 0.2, margins = c(2, 2), Colv = NA, scale='none')

# scaled
heatmap(t(imputed.crc.prot), cexRow = 0.3, cexCol = 0.2, margins = c(2, 2), Colv = NA)

heatmap(t(imputed.crc.prot), cexRow = 0.3, cexCol = 0.2, margins = c(2, 2), Colv = NA,
        col=my.colors)
```

### 4.2.3 Color bar for group information
```{r}
unique(imputed.crc.anno$Sub_group)
myColor <- rep("blue", nrow(imputed.crc.prot))
myColor[imputed.crc.anno$Sub_group == "CRC"] <- "red" 
myColor[imputed.crc.anno$Sub_group == "Benign"] <- "yellow" 
```

### 4.2.4 Standardized in a row (a protein)
quantile-spaced breaks in positive and negative directions 

```{r}
# change color code: protein-standardized expressions
# quantile-spaced breaks in positive and negative directions
imputed_CRC_prot_centerScale <-  ( t(imputed.crc.prot) - rowMeans(t(imputed.crc.prot)) ) / rowSds(t(imputed.crc.prot))

exprs_brk <- c( quantile(imputed_CRC_prot_centerScale[imputed_CRC_prot_centerScale < 0], probs=seq(0, 1, length=10)), 0, quantile(imputed_CRC_prot_centerScale[imputed_CRC_prot_centerScale > 0], seq(0, 1, length=10)))
colors <- colorRampPalette(c("darkblue", "white", "darkred"))(length(exprs_brk)-1)

heatmap(t(imputed.crc.prot), 
        cexRow = 0.3, cexCol = 0.2, margins = c(2, 2), Rowv = NA,
        col = colors, breaks = exprs_brk, ColSideColors=myColor)
```

### 4.2.5 Different distance and clustering

```{r, message=FALSE}
# can change method for distance calculation
col_distance <- dist(imputed.crc.prot, method = "euclidean")
# can change clustering method
col_cluster <- hclust(col_distance, method = "ward.D")

heatmap(t(imputed.crc.prot),
        cexRow = 0.3, cexCol = 0.2, margins = c(2, 2), 
        col = colors, 
        breaks = exprs_brk, 
        ColSideColors = myColor,
        Rowv = NA, 
        Colv = as.dendrogram(col_cluster)) 
```

